---
title: "Preprocessing"
author: "AGC"
date: "22 2 2022"
output: html_document
---

```{r analysis parameters, include=T}

threshold = 0.05
model = "~1+Group+Timepoint+Group*Timepoint"

```



```{r setup, include=FALSE}
home = getwd()

output=paste0(home,"/output/")
data=paste0(home, "/data/")

source(paste0(home,"/code/custom_functions.R"))

```

```{r load libraries, include=FALSE}
library(DESeq2)
library(tidyverse)
library(compareGroups)
library(limma)
library(lme4)
library(lm.beta)
library(viridis)
library(ggplot2)
```

```{r load data, include=FALSE}
counts=read.csv(paste0(data,"miRNAsRPMTable.csv"))
counts <- counts %>% column_to_rownames("customerID")
samplemeta <- read.csv(paste0(data,"SampleSheet.csv"))
samplemeta$customerID=paste0("X", samplemeta$customerID)
samplemeta <- samplemeta %>% column_to_rownames("customerID")
samplemeta$Group <- relevel(as.factor(samplemeta$Group), ref="Control")
samplemeta$Timepoint <- factor(samplemeta$Timepoint, ordered=T, 
                               levels=c("Baseline","d1","d7",
                                        "d14","End","postseizure"))

model = "~1+Group+Timepoint+Group*Timepoint"
table(samplemeta$Group,samplemeta$Timepoint)

drop_postseizure_idx <- samplemeta$Timepoint!="postseizure"
samplemeta <- samplemeta[drop_postseizure_idx, ]
samplemeta$Timepoint = droplevels(samplemeta$Timepoint)
counts <- counts[,rownames(samplemeta)]

```


```{r basic QC, fig.height=15, fig.width=15, echo=F}
log2_cpm = log2(counts+1)

varsset=apply(log2_cpm, 1, var)

limit=428
cpm.sel.trans = t(log2_cpm[order(varsset,decreasing = T)[1:limit],])

distance = dist(cpm.sel.trans)

sampleDistMatrix <- as.matrix(distance)

#colors for plotting heatmap
colors=viridis(255)

groupcol = Dark8[1:nlevels(samplemeta$Group)]
names(groupcol) = levels(samplemeta$Group)

timepointcol = colorRampPalette(c("lightblue", "dodgerblue4"))(nlevels(samplemeta$Timepoint))[c(1:nlevels(samplemeta$Timepoint))]
names(timepointcol) = levels(samplemeta$Timepoint)

ann_colors = list(
  Group = groupcol, 
  Timepoint = timepointcol
)

labels = samplemeta[,c("Group","Timepoint")] %>%  
  mutate_all(as.character) %>% as.data.frame()

rownames(labels)=rownames(samplemeta)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = distance,
         clustering_distance_cols = distance,
         clustering_method = "ward.D2",
         scale ="none",
         show_rownames=F, show_colnames = F,
         legend=T,
         border_color = NA, 
         annotation_row = labels,
         annotation_col = labels,
         annotation_colors = ann_colors,
         col = colors, 
         main = "D62 Distances normalized log2 counts")

```


## Statistical Modelling over time 

Time points where modeled as interaction term with Group status to identify which miRNAs where differential regulated over time in dependence of Group status. 

Time was modeled as polynomial function of the fourth degree to capture linear, quadratic, cubic and ^4 effects. 




```{r modelling, include=T, echo =F}
designform = as.formula(model)
design = model.matrix( designform, samplemeta)
fitDupCor <- lmFit(log2_cpm, design)
fit<- eBayes(fitDupCor)

lin <-  topTable(fit,n=dim(fit)[1], coef="GroupEpilepsy:Timepoint.L")
colnames(lin) = paste0("T^1xGroup_",colnames(lin))
squ  <- topTable(fit,n=dim(fit)[1], coef="GroupEpilepsy:Timepoint.Q")
colnames(squ) = paste0("T^2xGroup_",colnames(squ))
cub <- topTable(fit,n=dim(fit)[1], coef="GroupEpilepsy:Timepoint.C")
colnames(cub) = paste0("T^3xGroup_",colnames(cub))
qua <- topTable(fit,n=dim(fit)[1], coef="GroupEpilepsy:Timepoint^4")
colnames(qua) = paste0("T^4xGroup_",colnames(qua))

results = cbind(lin,
                squ[rownames(lin),],
                cub[rownames(lin),],
                qua[rownames(lin),])



adjsig = rowSums(results[,grepl("adj",colnames(results))]<threshold)>0
sig_adjusted=sum(rowSums(results[,grepl("adj",colnames(results))]<threshold)>0)

nomsig=rowSums(results[,grepl("P.Value",colnames(results))]<threshold)>0
sig_nom=sum(nomsig)
```

### Significant hits

In total `r sig_adjusted` miRNAs where siginificant after correction for multiple testing and a total of `r sig_nom` mir RNAs where significnat at an uncorrected p-value threshold of `r threshold`

### Nominal significnt hits 

```{r show modelling, include=T, echo =F}
display_tab(results[nomsig,grepl("P.Value",colnames(results))])

```

### unsorted heatmap of significnt hits 

```{r visualize, fig.height=15, fig.width=15, include=T, echo=F}
cpm.sel.trans<-t(log2_cpm[nomsig, order(samplemeta$Group, samplemeta$Timepoint)])
distance = dist(cpm.sel.trans)
sampleDistMatrix <- as.matrix(distance)

pheatmap(sampleDistMatrix,
         cluster_rows=F, 
         cluster_cols = F,
         scale ="none",
         show_rownames=F, show_colnames = F,
         legend=T,
         border_color = NA, 
         annotation_row = labels,
         annotation_col = labels,
         annotation_colors = ann_colors,
         col = colors, 
         main = "D62 Distances normalized log2 counts")

```

### time line plots of significnt hits 

```{r line-plots, fig.height=24, fig.width=8, include=T, echo=F}

plotdata <- cbind(samplemeta, t(log2_cpm[nomsig,]) )


plotdata <- plotdata %>% reshape2::melt(., direction = "long",
                            value.name = "counts",
                            variable.name="gene",
                            measure.vars = rownames(log2_cpm)[which(nomsig)])

p<-ggplot(plotdata, aes(x=Timepoint, y=counts, group=Group, col=Group)) +
  geom_smooth()+geom_jitter(width=0.05)+facet_wrap(vars(gene), ncol=4)+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p

```

